---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
library(tidyverse)
# Download the relevant weekly data
tuesdata <- tidytuesdayR::tt_load('2020-09-22')

# Selecting expeditions data
expeditions <- tuesdata$expeditions

# Selecting the 15 most common peaks
top_peaks <- expeditions %>%
  mutate(
    peak_lump = forcats::fct_lump(peak_name, n = 15)
  ) %>%
  filter(peak_lump != "Other")

# Sort by expedition number
top_peaks <- top_peaks %>%
  mutate(
    peak_lump = forcats::fct_reorder(peak_lump, peak_lump, .fun = length)
  )

# Preparation of the figure
ggplot(top_peaks, aes(x = peak_lump, fill = peak_lump)) +
  geom_bar() +
  coord_flip() +
  scale_fill_viridis_d(option = "D") +
  theme_light() +
  labs(
    x = "Peak",
    y = "Number of Expeditions",
    title = "Top 15 Most Frequent Peaks in Expeditions (TidyTuesday 2020-09-22)"
  ) +
  theme(legend.position = "none")
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}

```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}

```
