---
title: "TidyTuesday 2025-03-11" - "Pixar films"
author: "Vikár Míra"
output: html_document
---

In this project, I examine data from Pixar films.  
My goal is to explore data patterns (EDA) and then build, compare, and evaluate at least two models.  
The data analysis is fully reproducible, and all code is available in Rmd.

# Load packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(readr)
```

```{r}
# Importing the Pixar dataset directly from GitHub

pixar_films <- readr::read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-03-11/pixar_films.csv"
)

public_response <- readr::read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-03-11/public_response.csv"
)
```
```{r}
# Pixar dataset head and glimpse
head(pixar_films)
head(public_response)
knitr::kable(head(pixar_films))
knitr::kable(head(public_response))
glimpse(pixar_films)
glimpse(public_response)
```
## 1. Data cleaning
- adatok betöltése
- alap takarítás
- változók magyarázata
- fontos lépések kommentelve

```{r}
#Examining variable types and missing data
glimpse(pixar_films)
summary(pixar_films)

glimpse(public_response)
summary(public_response)
```

```{r}
#exemination of missing values
public_response %>%
  filter(is.na(critics_choice) | is.na(rotten_tomatoes))
```

Data Overview and Initial Cleaning Steps
Before starting the analysis, I examined the dataset to understand its structure and identify potential data quality issues. The dataset contains 24 Pixar films and includes the following variables:

film (character) – The title of the movie.

rotten_tomatoes (numeric) – Percentage score from Rotten Tomatoes.

metacritic (numeric) – Metacritic critic score (0–100).

cinema_score (character) – Audience grade (A, A+, etc.).

critics_choice (numeric) – Critics’ Choice rating (0–100).

Summary of Key Findings
A first look using glimpse() and summary() revealed the following:

All variables were loaded correctly with the expected data types.

cinema_score is stored as a character variable, not numeric, because it contains letter-based grades (e.g., A, A+).

Some variables include missing values (NA):

critics_choice: 3 missing values

metacritic: 1 missing value

cinema_score: 1 missing value

I also inspected rows containing missing data. These came exclusively from the film “Luca” and two early films where Critics Choice ratings were not available.

Next Step: Data Cleaning Plan
Since the dataset is small and missing values are minimal, carefully handling NAs is important. I will:
Convert cinema_score to an ordered factor
Dealing with missing values.

```{r}
# Convert cinema_score to an ordered factor
public_response <- public_response %>%
  mutate(
    cinema_score = factor(
      cinema_score,
      levels = c("A-", "A", "A+"),   # adjust if more categories appear
      ordered = TRUE
    )
  )

```

## 2. Exploratory Data Analysis (EDA)
- legalább **két ember számára érthető, jól feliratozott plot**
- rövid értelmezés a grafikák alatt
- táblázatok és statisztikák

```{r}


```

## 3. Modellek építése
- legalább **két modell** ugyanarra az outcome változóra
  - pl. lineáris modell, logisztikus, random forest, stb.
- modell-összehasonlítás
  - AIC, RMSE, accuracy stb.
- rövid konklúzió: melyik jobb és miért?

```{r}


```

## 4. Modellek értékelése
- **assumption check + residual diagnostics**
  - pl. QQ plot, residuals vs fitted plot
- értelmezés: mi látszik, rendben vannak-e a feltételek?

```{r}


```

## 5. Következtetések / Report
- összefoglalás emberi nyelven
- mik a tanulságok?
- mit találtál érdekesnek?

```{r}


```

## 6. Kreatív rész (ha szeretnél)
- extra vizualizációk
- extra kérdések
- saját ötletek

```{r}


```
